---
title: "Creating networks from survey data"
author: "Elliot Meador"
date: "28 June 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(igraph)
library(tidyverse)
library(RColorBrewer)
library(visNetwork)
```

 # Introduction

This code is from a talk I gave at EdinR back in March on using social network analysis with survey data in R. I used some research data in that talk, but here I'll show how networks can be created with any survey-style research. 

First, let's create a network object that looks and behaves like a network of people interacting with another. For this, we use the `igraph` package and the `erdos.renyi.game` function to create a graph that resembles people. We then grab the edge list in dataframe format to use in the subsequent steps. 
```{r}
# Realistic graph from scratch ---------------
set.seed(123456) # for replicabiltiy 
(g <- erdos.renyi.game(75, .025, loops = F)) # this creates a graph object

g_1 <- induced_subgraph(g, which(degree(g) >= 1)) # this gets rid of any unconnected nodes

V(g_1)$name <- 1:length(V(g_1)) # assign a name to the vertices as an attribute
```

Let's take a look at the graph, called `g_1`. This can be done using `R`'s base `plot` function

```{r}
plot(g_1)
```

`igraph` is best for analysing 'under-the-hood' of graphs, though its plotting functions are very powerful. We'll use `visNetwork` to visualise the graph. But first we need to extract some node attributes. 

Let's use `map`  from the `purrr` package to grab the edges and nodes, both as `tibbles`. The edges `tibble` will be our working dataset and it mimics how data in surveys may be structured. 
```{r}

g_1_ls <- map(get.data.frame(g_1, 'both'), as_tibble)

edges <- g_1_ls$edges

big_cliques <- cluster_walktrap(g_1) %>% # this bit will allow the gender variable to be
  membership() %>%                       # non-random in a way that is evident in the graph
  as.list() %>% 
  flatten_df() %>% 
  gather(respondent,community) %>% 
  mutate(respondent = as.numeric(respondent))

```

A common approach to conducting network analysis with surveys is to have one variable designated as a 'from' variable and another as a 'to' variable. Sometimes these columns are referred to as 'source' and 'target', but they represent the same thing. 'From' is the respondent; and 'to' is the person to whom the respondent goes to for information. Instead of removing them from the graph we will assign them a special attribute so we can recognise them in the graph.

```{r}

# Example data ---------------------

example_df <- edges %>%
  purrr::set_names('respondent', 'contact_for_info') %>% # create survey style attributes
  left_join(big_cliques) %>%                             # and graph from data frame.
  mutate(resp_gender = ifelse(community %in% c(1,2,3), 
                              'Female', 'Male')) 

```

We use the `graph_from_data_frame` from `igraph` to make a network. `graph_from_data_frame` assumes the first two columns are for constructing the network and assigns the remaining columns as edge attributes. Then we use `dplyr` to merge and mutate the existing attribute dataframe.

```{r}
example_graph <- graph_from_data_frame(example_df[1:2], directed = T)

```

We now need to extract all nodes from the graph and use this to create a node attribute dataframe. It is highly likely that there will be missing data in networks created from surveys if the `from` variable is left open-ended. Leaving the `from` variable as open-ended (i.e. 'Name one person you go to for information on recylcling . . .') is often neccesary as possible answers are not know to the researcher prior to running the survey.


```{r}
# get the nodes dataset from the graph object
nodes <- get.data.frame(example_graph, 'vertices') %>%
  as_tibble() %>%
  purrr::set_names('respondent') %>%
  mutate_all(as.numeric)

# get the original survey data and drop the 'to' column
join_df <- example_df %>% select(-contact_for_info)

# join the nodes data and survey data and change the NA values to unknown
node_attributes <- left_join(nodes, join_df) %>% 
  distinct() %>%
  mutate(resp_gender = ifelse(is.na(.$resp_gender), 'Unknown', resp_gender))

```
We now have our graph object and a node attributes dataframe we can use to 'decorate' the graph with. First, lets create a color function to assign colors and a rescale function to resize the nodes.

```{r}

#create colors function using RColorBrewer
Pastel2 <- colorRampPalette(sample(brewer.pal(8, 'Pastel2'),4))

# function to rescale size
rescale <- function(nchar, low, high) {
  min_d <- min(nchar)
  max_d <- max(nchar)
  rscl <- ((high - low) * (nchar - min_d)) / (max_d - min_d) + low
  rscl
}

# use dplyr to color the unique genders and add a column of colors
# 
# the degree function is from igraph and grabs each nodes degree
node_attributes <- node_attributes %>%
  select(resp_gender) %>%
  distinct() %>%
  mutate(color = Pastel2(nrow(.))) %>%
  right_join(node_attributes) %>%
  select(respondent, gender = resp_gender, color) %>%
  mutate(size = rescale(degree(g_1), 10, 50), # sizing graph nodes according to their degree
         size = ifelse(gender == 'Unknown', 0, size))


# We use igraphs bespoke notation to 'decorate' the graph with attributes
V(g_1)$color <- node_attributes$color
V(g_1)$size <- node_attributes$size
V(g_1)$gender <- node_attributes$gender

```

Now that the graph is decorated with the appropriate attributes, we will use the `visNetwork` package to make an interactive graph. `visNetwork` has functions that allow it to work directly with `igraph` graph objects. `visNetwork` uses dataframes to graph, so first we turn the graph into the appropriate dataframe and then plot.
```{r}
# Visualise using visNetwork package --------------------------------------

vis_dfs <- toVisNetworkData(g_1)


visNetwork(
  edges = vis_dfs$edges,
  nodes = vis_dfs$nodes,
  main  = 'Network example',
  submain = 'Nodes are sized by degree and\ncoloured by respondent gender'
) %>%
  visEdges(arrows = "from", dashes = T) %>%
  visInteraction(navigationButtons = TRUE) %>%
  visOptions(highlightNearest = T, selectedBy = 'gender')


```

